Liveness-Based Garbage Collection for Lazy Languages

### 2015-05-07 paper/paper.tex
changed liveness

For  the  case  of a  user-function  call  we  use a  third  parameter
$\Lfonly$  that  represents the  summaries  of  all  functions in  the
program.  $\Lfonly_f$  (the  component  of $\Lfonly$  for  a  specific
function $f$),  expresses how the demand  $\sigma$ on the  result of a
call to $f$ is transformed into  the liveness of its parameters at the
beginning of the call.  $\Lfonly$  is determined by the judgement form
$\mathit{Prog} \len \Lfonly$ using inference rule ({\sc live-define}).
This  rule  describes the  fixed-point  property  to  be satisfied  by
$\Lfonly$, namely, the demand transformation assumed for each function
in  the  program should  be  the  same  as the  demand  transformation
calculated    from   their    bodies.     As   we    shall   see    in
Section~\ref{sec:grammar-formulation},  the least  solution satisfying
the  rule  is obtained  by  converting  the  rule into  an  equivalent
grammar (actually the language recognized by the grammar is the least solution).  The least solution also  ensures that the greatest amount of
garbage is collected.

We next  describe the function $\mathcal{L}$  that propagates liveness
across  expressions.   This brings  us  to  the  second way  in  which
liveness analysis for lazy languages differ from eager languages---the
point  where  an  expression   gets  evaluated  cannot  be  statically
determined.  To  deal with this, we introduce an analysis which is
flow-insensitive but path-sensitive\comment{Checked the terminology, path-sensitive means certain prooperties holding along a path only, http://en.wikipedia.org/wiki/Data-flow\_analysis} into
our  analysis.

\begin{figure*}[t!]
\begin{picture}(100,130)(0, -50)
  \begin{tabular}{cc}
    \begin{boxedminipage}{.51\textwidth}
      {\sf
        \small
        \renewcommand{\arraystretch}{1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\length~\xl)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\ $\psi_1\!\!:\,$ \px
            \UNL{4} \hspace*{1.15cm} $\psi_2\!\!:\, (\RETURN ~0)$
            \UNL{4} \hspace*{1.15cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
            \UNL{5} \hspace*{1.15cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
            \UNL{6} \hspace*{1.15cm} $\pi_7\!\!:\,
            $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
            \UNL{7} \hspace*{1.15cm} $\psi_3\!\!:\, (\RETURN ~\pz)$))))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $\pi_\mainpgm\!\!:\, \pi_8\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
	  \UNL{2}   \hspace*{.78cm}   $\pi_9\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
          \UNL{3}   \hspace*{.78cm}    $\pi_{10}\!\!:\,
          $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN
          \UNL{4}  \hspace*{.78cm}  $\psi_{4}\!\!:\,$ (\RETURN\ \pw)))))))
\end{uprogram}}
        } 
    \end{boxedminipage}
    & 
    \begin{boxedminipage}{.49\textwidth}
      {\sf
        \small
        \begin{tabular}{lc}
        $\cal{M_{\mbox{\tiny \length}}} = $ 
            \begin{tabular}{lc}            
              $[$ \\
              $\psi_1 \mapsto \lbrace \px.\sigma_{\length}, \xl.\cbar\sigma_{\length} \rbrace ,$ \\ 
              $\psi_2 \mapsto \lbrace \rbrace , $\\
              $\psi_3 \mapsto \lbrace \pz.\sigma_{\length}, \py.\cbar\sigma_{\length}, $\\
              $\pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}} \rbrace $  \\
              $]$
            \end{tabular} \\
            $\cal{P_{\mbox{\tiny \length}}} = $
              \begin{tabular}{lc}
                $[$ \\
                $\pi_{\mbox{1}} \mapsto \lbrace \psi_1, \psi_2, \psi_3 \rbrace, $ \
                $ \pi_{\mbox{2}} \mapsto \lbrace \psi_1, \psi_2, \psi_3 \rbrace, $ \\
                $ \psi_{\mbox{1}} \mapsto \lbrace \psi_1 \rbrace, $ \
                $ \psi_{\mbox{2}} \mapsto \lbrace \psi_2 \rbrace, $ \
                $ \pi_{\mbox{5}} \mapsto \lbrace \psi_3 \rbrace, $ \\
                $ \pi_{\mbox{6}} \mapsto \lbrace \psi_3 \rbrace, $ \
                $ \pi_{\mbox{7}} \mapsto \lbrace \psi_3 \rbrace $ \\
                $ \psi_{\mbox{3}} \mapsto \lbrace \psi_3 \rbrace $ \\
                $]$
              \end{tabular}\\
              $\cal{M_{\mbox{\tiny \main}}} = $ 
              \begin{tabular}{lc}            
                $[$
                  $\psi_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace $ \
                  $]$
              \end{tabular} \\
              $\cal{P_{\mbox{\tiny \main}}} = $
                \begin{tabular}{lc}
                  $[$ \
                  $\pi_{\mbox{8}} \mapsto \lbrace \psi_4 \rbrace, $ \
                  $ \pi_{\mbox{9}} \mapsto \lbrace \psi_4 \rbrace, $ \
                  $ \pi_{\mbox{10}} \mapsto \lbrace \psi_4 \rbrace, $ \
                  $ \psi_{\mbox{4}} \mapsto \lbrace \psi_4 \rbrace, $ \
                  $ \ldots  $ \
                  $]$  
                \end{tabular}
        \end{tabular}
      }
    \end{boxedminipage}
    \\
    (a) Example program.&
    \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
        (b) {\em e-path} to liveness environment map \\
        (c) Prog pt to {\em e-paths } map
    \end{tabular}}
  \end{tabular}
\end{picture}
\kern -3ex
\vspace{0.5in}
\caption{Example Program and its liveness maps.}\label{fig:mot-example2}    
\end{figure*}

@@

Consider  an analysis of  the body  of $\length$  with a  demand $\sigma$.
Clearly, this is also the demand on the return value $z$ giving the
liveness  at  the  program point  $\psi_3$ as $\{\pz.\sigma\}$.   
The  liveness at $\psi_3$  is  propagated  backwards  as  in  traditional
liveness analysis.

Since   \pz\  is  $(1 + ~\py)$,  the   liveness  of   $\sigma$  for
\pz\ translates  into a demand  of $\sigma$ on $(1+~\py)$,  which in
turn  generates  a  liveness  of $\{\py.\epsilon\}$  at
$\pi_7$.  The liveness of \py\ should be killed at $\pi_6$ since it is
beyond the scope of the definition of \py.  However, this does not cause any imprecision
in the way of the GC marking  more cells as live, since \py\ is yet to
come into existence at $\pi_6$  and its liveness will not be consulted
by the garbage collector at this point.
In an eager language, \py\ would not have been live at $\psi_3$, the
point of  its last  use. In a  lazy language, however,  the definition
$\pz  \leftarrow (1+~\py)$,  does  not result  in  an evaluation  of
$(1+~\py)$; instead  a closure is created.   The evaluation actually
takes place while evaluating  $(\RETURN~\pz)$. Thus it is necessary to
record the liveness  of $\py$ at the program  point $\psi_3$, indeed at
every program  point from $\psi_3$ up  to the beginning  of the program
(since  we do  not kill  liveness).  We  thus regard  the body  of the
function as  a tree and consider paths  from the root of  this tree to
program  points  which  trigger  a evaluation\footnote{Program points
  which trigger evaluation are labelled with $\psi$}---\SIF\  conditions  and
\RETURN\  expressions.  We  call such  paths {\em
  e-paths}. \comment{Update the e-path labels}  For the
function \length, there are three such paths $\psi_1 = \pi_1, \pi_2, \pi_3,
\pi_4,  \pi_5$, $\psi_2  =  \pi_1, \pi_2,  \pi_3,  \pi_4, \pi_6,  \pi_7,
\pi_8$, and $\psi_3 = \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.

The analysis  consists of  two maps: The  map $\cal{M}$ maps  a e-path to a common liveness environment, and a second map $\cal{P}$ maps
a program  point to the set of  e-paths on which  the point lies.
If a  program  point happens to fall  on more than one  such path, then
the  liveness environment of  the program  point is  the variable-wise
union of  the liveness environments  of the individual paths.   In the
example program, ${\cal M}(\psi_1)= \ldots$, ${\cal M}(\psi_2)=\ldots$ and
${\cal M}(\psi_3)=  \ldots$.  Since  ${\cal P}(pi_4)$ is  $\lbrace \psi_1,
\psi_2, \psi_3 \rbrace$, the liveness  of $pi_4$ is \ldots.  In summary,
our propagation of liveness is flow-insensitive but path-sensitive.
In the sequel
we  shall  describe  how  to  obtain $\cal{M}$.   The  formulation  of
$\cal{P}$ is easy and we do not elaborate it any further.

>>>>

For  the  case  of a  user-function  call  we  use a  third  parameter
$\Lfonly$  that  represents the  summaries  of  all  functions in  the
program.   $\Lfonly_{\mathit f}$  (the  component of  $\Lfonly$ for  a
specific  function $f$),  expresses  how the  demand  $\sigma$ on  the
result  of a  call to  $f$  is transformed  into the  liveness of  its
parameters at the  beginning of the call.  $\Lfonly$  is determined by
the judgement  form $\mathit{Prog} \len \Lfonly$  using inference rule
({\sc live-define}).  This rule  describes the fixed-point property to
be satisfied  by $\Lfonly$, namely, the  demand transformation assumed
for each  function in  the program  should be the  same as  the demand
transformation  calculated from  their  bodies.  As  we  shall see  in
Section~\ref{sec:grammar-formulation},  we  convert  the rule  into  a
grammar and  the least  solution satisfying the  rule is  the language
generated by this  grammar.  The least solution also  ensures the safe
collection of the greatest amount of garbage.

We next  describe the function $\mathcal{L}$  that propagates liveness
across  expressions.   This brings  us  to  the  second way  in  which
liveness analysis for lazy languages differ from eager languages---the
point  where  an  expression   gets  evaluated  cannot  be  statically
determined.   %% To deal  with this,  we introduce  an analysis  which is
%% flow-insensitive  but path-sensitive\comment{Checked  the terminology,
%%   path-sensitive means certain prooperties  holding along a path only,
%%   http://en.wikipedia.org/wiki/Data-flow\_analysis} into our analysis.
To see  the consequences of this, consider  an analysis  of the body  of $\length$  with a
demand $\sigma$.  Clearly, this is also the demand on the return value
$z$   giving  the   liveness  at   the  program   point   $\psi_3$  as
$\{\pz.\sigma\}$.  The liveness at $\psi_3$ is propagated backwards as
in traditional liveness analysis.

\begin{figure*}[t!]
\begin{picture}(100,130)(0, -50)
  \begin{tabular}{cc}
    \begin{boxedminipage}{.51\textwidth}
      {\sf
        \small
        \renewcommand{\arraystretch}{1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\length~\xl)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\ $\psi_1\!\!:\,$ \px
            \UNL{4} \hspace*{.27cm} $\psi_2\!\!:\, (\RETURN ~0)$
            \UNL{4} \hspace*{.29cm}    $\pi_3\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
            \UNL{5} \hspace*{.34cm}   $\pi_4\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
            \UNL{6} \hspace*{.34cm} $\pi_5\!\!:\,
            $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
            \UNL{7} \hspace*{.34cm} $\psi_3\!\!:\, (\RETURN ~\pz)$))))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $\pi_\mainpgm\!\!:\, \pi_6\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
	  \UNL{2}   \hspace*{.78cm}   $\pi_7\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
          \UNL{3}   \hspace*{.78cm}    $\pi_8\!\!:\,
          $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN
          \UNL{4}  \hspace*{.78cm}  $\psi_{4}\!\!:\,$ (\RETURN\ \pw)))))))
\end{uprogram}}
        } 
    \end{boxedminipage}
    & 
    \begin{boxedminipage}{.49\textwidth}
      {
        \small
        \begin{tabular}{lc}
        $\cal{M_{\mbox{\tiny \length}}} = $ 
            \begin{tabular}{lc}            
              $[$ 
              $ep_1 \mapsto \lbrace \px.\sigma_{\length}, \xl.\cbar\sigma_{\length} \rbrace ,~ep_2 \mapsto \lbrace \rbrace , $\\
              $~~ep_3 \mapsto \lbrace \pz.\sigma_{\length},
                \py.\cbar\sigma_{\length}, \pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, $\\
              $~~~~~~~~~~~~~~\;\; \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}} \rbrace ]$
            \end{tabular} \\
            $\cal{P_{\mbox{\tiny \length}}} = $
              \begin{tabular}{lc}
                $[~\pi_1 \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \
                $ \pi_2 \mapsto \lbrace ep_1, ep_2, ep_3
                  \rbrace, $\\
                $~~\psi_1 \mapsto \lbrace ep_1 \rbrace, $ \
                $ \psi_2 \mapsto \lbrace ep_2 \rbrace, $ \
                $ \pi_5 \mapsto \lbrace ep_3 \rbrace, $ \\
                $~~\pi_6 \mapsto \lbrace ep_3 \rbrace, $ \
                $ \pi_7 \mapsto \lbrace ep_3 \rbrace,~ \psi_3 \mapsto \lbrace ep_3 \rbrace\rbrack$
              \end{tabular}\\
              $\cal{M_{\mbox{\tiny \main}}} = $ 
              \begin{tabular}{lc}            
                $[$
                  $ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace $ \
                  $]$
              \end{tabular} \\
              $\cal{P_{\mbox{\tiny \main}}} = $
                \begin{tabular}{lc}
                  $~\lbrack$~$\pi_6 \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \pi_7 \mapsto \lbrace ep_4 \rbrace, $ \
                  $\pi_8 \mapsto \lbrace ep_4 \rbrace, $ \\
                  $~~~ep_4 \mapsto \lbrace ep_4 \rbrace,\ldots~\rbrack$  
                \end{tabular}
        \end{tabular}
      } 
    \end{boxedminipage}
    \\
    (a) Example program.&
    \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
        (b) {\em e-path} to liveness environment map \\
        (c) Prog pt to {\em e-paths } map
    \end{tabular}}
  \end{tabular}
\end{picture}
\kern -3ex
\vspace{0.5in}
\caption{Example Program and its liveness maps.}\label{fig:mot-example2}    
\end{figure*}

@@

Since   \pz\  is  $(1 + ~\py)$,  the   liveness  of   $\sigma$  for
\pz\ translates  into a demand  of $\sigma$ on $(1+~\py)$,  which in
turn  generates  a  liveness  of $\{\py.\epsilon\}$  at
$\pi_7$.  The liveness of \py\ should be killed at $\pi_6$ since it is
beyond the scope of the definition of \py.  However, this does not cause any imprecision
in the way of the GC marking  more cells as live, since \py\ is yet to
come into existence at $\pi_6$  and its liveness will not be consulted
by the garbage collector at this point.
In an eager language, \py\ would not have been live at $\psi_3$, the
point of  its last  use. In a  lazy language, however,  the definition
$\pz  \leftarrow (1+~\py)$,  does  not result  in  an evaluation  of
$(1+~\py)$; instead  a closure is created.   The evaluation actually
takes place while evaluating  $(\RETURN~\pz)$. Thus it is necessary to
record the liveness  of $\py$ at the program  point $\psi_3$, indeed at
every program  point from $\psi_3$ up  to the beginning  of the program
(since  we do  not kill  liveness).  We  thus regard  the body  of the
function as  a tree and consider  paths  from the root of  this tree to
program  points  which  trigger  a evaluation\footnote{Program points
  which trigger evaluation are labelled with $\psi$}---\SIF\  conditions  and
\RETURN\  expressions.  We  call such  paths {\em
  e-paths}. \comment{Update the e-path labels}  For the
function \length, there are three such paths $ep_1 = \pi_1, \pi_2, \psi_1$,
$ep_2 = \pi_1,  \pi_2, \psi_2$ and  $ep_3  =  \pi_1, \pi_2,  \pi_3,
\pi_4, \pi_5,  \psi_3$.

The analysis consists of two maps:  The map $\cal{M}$ maps a e-path to
a liveness environment arising out  of all uses of variables along the
path, and  a second map $\cal{P}$ maps  a program point to  the set of
e-paths sharing the point.  If a program point happens to fall on more
than one such path, then the liveness environment of the program point
is  the  variable-wise  union  of  the liveness  environments  of  the
individual paths.  In the example program, ${\cal M}(\psi_1)= \ldots$,
${\cal  M}(\psi_2)=\ldots$  and  ${\cal  M}(\psi_3)=  \ldots$.   Since
${\cal  P}(pi_4)$ is  $\lbrace  \psi_1, \psi_2,  \psi_3 \rbrace$,  the
liveness of $pi_4$ is \ldots.  In summary, our propagation of liveness
is  flow-insensitive  but  path-sensitive.   In the  sequel  we  shall
describe  how to obtain  $\cal{M}$.  The  formulation of  $\cal{P}$ is
easy and we do not elaborate it any further.
